{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-10 thesis/overview.tex
polish background chapter; was pretty stale after being copied from the proposal

This chapter introduces the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.

\section{Concurrency}

\subsection{The Basics}

@@

\subsection{Transactional Memory}
\label{sec:overview-tm}

@@

Some model checkers explicitly store the set of visited program states as a means of identifying equivalent interleavings \cite{spin}.
From the perspective of such tools, state spaces such as these wherein equivalent states may be reached by multiple paths
are represented as a directed acyclic graph (DAG) instead of as a tree.
This approach is called \revisionminor{{\em stateful} model checking \cite{spin}}.
This thesis focuses on {\em stateless} model checking (and execution trees, not DAGs),
which instead analyzes the sequence of execution events to avoid a prohibitive memory footprint.
Henceforth ``stateless model checking'' will be abbreviated simply as ``model checking'' for brevity.
Also, the term ``state space'' was originally coined to refer to the stateful approach's emphasis on the DAG's nodes
(i.e., program states);
while stateless checkers emphasize the tree's branches (i.e., execution sequences) instead,
I will continue to use ``state space'' for consistency with prior work.

@@

{\bf Abstractly speaking:}
Let {\em independent transitions} denote a pair of executions of two threads, each from one preemption point to the next,
in which there are no read/write or write/write access pairs to the same memory between threads.
DPOR reduces a state space, originally exponentially-sized in the number of thread transitions,
to an equivalent one
(i.e., testing which suffices to check all program behaviours that could arise in the original state space)
exponentially-sized in the number of {\em dependent} thread transitions.
%More intuitively, if two thread transitions between preemption points do not conflict on any shared resource access,
%reordering them produces an equivalent interleaving, i.e., the same program behaviour.
More technically, it identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
and tests at least one execution from each equivalence class.

{\bf Concretely speaking:}
\Cref{fig:dpor} highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
and each interleaving has a respective ``subtree'' (i.e., possible interleavings given the fixed execution prefix leading up to it).
The specifics of execution before the thread 1/thread 2 sequence,
other possible threads to run instead of threads 1 or 2,
and what logic the program executes in those subtrees
are all presumably arbitrary.
In these two highlighted branches,
if the transitions of threads 1 and 2 are {\em independent},
%if the operations performed by threads 1 and 2 are independent
%(i.e., no write/read or write/write access pairs to the same memory),
DPOR deduces that the subsequent program states (indicated by the red arrow) are equivalent.
Thence, only one of the two interleavings and its respective subtree needs to be executed
in order to check all possible program states.
\cref{sec:landslide-dpor} explains how DPOR implements such a deduction in more detail.

@@

\begin{figure}[t]
	\begin{center}
	\includegraphics[width=0.6\textwidth]{dpor-new-colors.pdf}
	\end{center}
	\caption[DPOR identifies independent transitions by different threads.]
		{DPOR identifies independent transitions by different threads which can commute
	without affecting program behaviour.
	Here, because each thread's local operations on {\tt tmp1}/{\tt tmp2} are not shared memory conflicts,
	the states marked with the red arrow are guaranteed identical. Hence, only one of the subtrees need be explored.}
	\label{fig:dpor}
\end{figure}

@@

\section{Data Race Analysis}
\label{sec:background-datarace}

@@

Although the interfaces are not mandated \revisionminor{as} they were in P2, all Pebbles kernels must necessarily contain the same abstract components. These include:
\begin{itemize}
	\item A round-robin scheduler, including context switching, timer handling, and runqueue management;
	\item Some approach to locking, often analogous to P2's concurrency primitives (henceforth referred to as ``kernel mutexes''),
		and some approach to blocking threads indefinitely;
	\item A virtual memory implementation \revisionminor{(page directories and tables, frame allocator, et cetera)};
	\item A program loader;
	\item Lifecycle management code for creation and destruction of \revisionminor{threads and tasks};
	\item Other miscellany such as a suite of fault handlers to ensure no user program can cause the kernel itself to crash.
\end{itemize}
Because any combination of system calls or fault handlers can be invoked by user programs simultaneously,
concurrency bugs can arise from the interaction of any subset of kernel components with each other.
The most common bugs students face arise from the interaction of some component with itself (e.g., concurrent invocations of {\tt new\_pages}/{\tt remove\_pages} in the same process),
or from the interaction between an exiting thread and some other thread trying to communicate with it ({\tt vanish} versus, well, anything else, really).
The most difficult concurrency problem in P3 is that of coordinating a parent and a child task that simultaneously exit:
when a task completes, live children and exited zombies must be handed off to the task's parent or to the {\tt init} process,
when the task's parent may itself be exiting;
meanwhile, threads in tasks that receive new children may need to be awakened from {\tt wait}.
Careless solutions to this problem are prone to data races or deadlocks.

@@

The Pintos kernel architecture \cite{pintos} is used at several universities, including Berkeley, Stanford, and the University of Chicago.
The Pintos basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader,
upon which students add more features in several projects.
Most relevant to this thesis, the basecode provides the following functions/libraries, among others:
\begin{itemize}
	\item Semaphores (the basic concurrency primitive, implemented using direct scheduler calls): {\tt sema\_up}, {\tt sema\_down}, {\tt sema\_try\_down};
	\item Locks (which wrap a semaphore initialized to 1), {\tt lock\_acquire}, {\tt lock\_\allowbreak{}release}, {\tt lock\_try\_acquire};
	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait}, {\tt cond\_\allowbreak{}signal}, {\tt cond\_broadcast}, with the same semantics as Pebbles P2 condvars;
	\item Basic round-robin scheduling facilities: {\tt thread\_block} (a kernel-level analogue to Pebbles's {\tt deschedule}), {\tt thread\_yield}
	\item Kernel thread lifecycle management, {\tt thread\_create} and {\tt thread\_exit}, including stack space memory management;
	\item Interrupt and fault handlers;
	\item A page allocator, {\tt palloc\_get\_page}, {\tt palloc\_get\_multiple}, {\tt palloc\_free\_page}, and {\tt palloc\_free\_multiple}
\end{itemize}
Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc}, string-formatting, printing, et cetera.

>>>>

\revisionminor{This chapter introduces the necessary background material on concurrency
(\cref{sec:overview-concurrency}),
stateless model checking,
(\cref{sec:overview-stateless}),
data-race analysis
(\cref{sec:background-datarace}),
and the relevant undergraduate operating systems classes
(\cref{sec:overview-edu}).
It concludes with a glossary of terminology
(\cref{sec:glossary})
that I hope the reader may find convenient as a quick reference
while reading later chapters as well.}

\section{Concurrency}
\label{sec:overview-concurrency}

\revision{This section presents an elementary introduction to
%the fundamentals of
concurrent programming.
I have taken care to make it accessible to readers with even only basic programming experience,
familiar with flow control, functions, memory, and objects at most,
although I cannot avoid the inherent difficulty of multiple layers of abstraction building upon one another.
For an explanation to be ``elementary''
means that it requires very little be known ahead of time to approach it,
not necessarily that it be easy
\cite{lost-lecture}.
%\footnote{``Elementary'' does not mean easy to understand.
%``Elementary'' means that very little is required to know ahead of time in order to understand it,
%except to have an infinite amount of intelligence
%\cite{lost-lecture}.}
}

\subsection{The basics}

@@

\subsection{Transactional memory}
\label{sec:overview-tm}

@@

Some model checkers explicitly store the set of visited program states as a means of identifying equivalent interleavings \cite{spin}.
From the perspective of such tools, state spaces such as these wherein equivalent states may be reached by multiple paths
are represented as a directed acyclic graph (DAG) instead of as a tree.
This approach is called \revisionminor{{\em stateful} model checking.} %\cite{spin}.
This thesis focuses on {\em stateless} model checking (and execution trees, not DAGs),
which instead analyzes the sequence of execution events to avoid a prohibitive memory footprint.
Henceforth ``stateless model checking'' will often be abbreviated simply as ``model checking'' for brevity.
Also, the term ``state space'' was originally coined to refer to the stateful approach's emphasis on the DAG's nodes
(i.e., program states);
while stateless checkers emphasize the tree's branches (i.e., execution sequences) instead,
I will continue to use ``state space'' for consistency with prior work.

@@

%{\bf Abstractly speaking:}
\revisionminor{To define it abstractly, let}
{\em independent transitions} denote a pair of executions of two threads, each from one preemption point to the next,
in which there are no read/write or write/write access pairs to the same memory between threads.
DPOR reduces a state space, originally exponentially-sized in the number of thread transitions,
to an equivalent one
(i.e., testing which suffices to check all program behaviours that could arise in the original state space)
exponentially-sized in the number of {\em dependent} thread transitions.
%More intuitively, if two thread transitions between preemption points do not conflict on any shared resource access,
%reordering them produces an equivalent interleaving, i.e., the same program behaviour.
More technically, it identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
and tests at least one execution from each equivalence class.

%{\bf Concretely speaking:}
\revisionminor{For a friendlier, more concrete example, consider \Cref{fig:dpor}, which}
highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
and each interleaving has a respective subtree (i.e., possible interleavings given the fixed execution prefix leading up to it).
\revisionminor{Any events executed}
before the thread 1/thread 2 sequence,
any other runnable threads besides these two, %threads 1 or 2,
and what logic the program executes in those subtrees
are all presumably arbitrary.
In these two highlighted branches,
if the transitions of threads 1 and 2 are {\em independent},
%if the operations performed by threads 1 and 2 are independent
%(i.e., no write/read or write/write access pairs to the same memory),
DPOR deduces that the subsequent program states (indicated by the red arrow) are equivalent.
Thence, only one of the two interleavings and its respective subtree needs to be executed
in order to check all possible program states.
\cref{sec:landslide-dpor} explains how DPOR implements such a deduction in more detail.

@@

\begin{figure}[t]
	\begin{center}
	\includegraphics[width=0.55\textwidth]{dpor-new-colors.pdf}
	\end{center}
	\caption[DPOR identifies independent transitions by different threads.]
		{DPOR identifies independent transitions by different threads which can commute
	without affecting program behaviour.
	Here, because each thread's local operations on {\tt tmp1}/{\tt tmp2} are not shared memory conflicts,
	the states marked with the red arrow are guaranteed identical. Hence, only one of the subtrees need be explored.}
	\label{fig:dpor}
\end{figure}

@@

\section{Data race analysis}
\label{sec:background-datarace}

@@

Although the interfaces are not mandated \revisionminor{as} they were in P2, all Pebbles kernels must necessarily contain the same abstract components. These include:
\begin{itemize}
	\item A round-robin scheduler, including context switching, timer handling, and runqueue management;
	\item Some approach to locking, often analogous to P2's concurrency primitives (henceforth referred to as ``kernel mutexes''),
		and some approach to blocking threads indefinitely;
	\item A virtual memory implementation \revisionminor{(page directories and tables, frame allocator, et cetera)};
	\item A program loader;
	\item Lifecycle management code for creation and destruction of \revisionminor{threads and tasks};
	\item Other miscellany such as a suite of fault handlers to ensure no user program can cause the kernel itself to crash.
\end{itemize}
Because user programs can invoke any combination of system calls or fault handlers simultaneously,
concurrency bugs can arise from the interaction of any subset of kernel components with each other.
The most common bugs students face arise from the interaction of some component with itself (e.g., concurrent invocations of {\tt new\_pages}/{\tt remove\_pages} in the same process),
or from the interaction between an exiting thread and some other thread trying to communicate with it ({\tt vanish} versus, well, anything else, really).
The most difficult concurrency problem in P3 is that of coordinating a parent and a child task that simultaneously exit:
when a task completes, live children and exited zombies must be handed off to the task's parent or to the {\tt init} process,
when the task's parent may itself be exiting;
meanwhile, threads in tasks that receive new children may need to be awakened from {\tt wait}.
Careless solutions to this problem are prone to data races or deadlocks.

@@

The Pintos kernel architecture \cite{pintos} is used at several universities, including the University of Chicago, Berkeley, and Stanford.
Its basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader,
upon which students add more features in several projects.
Most relevant to this thesis, the basecode provides the following functions/libraries, among others:
\begin{itemize}
	\item Semaphores (the basic concurrency primitive, implemented using direct scheduler calls): {\tt sema\_up}, {\tt sema\_down}, {\tt sema\_try\_down};
	\item Locks (which wrap a semaphore initialized to 1), {\tt lock\_acquire}, {\tt lock\_\allowbreak{}release}, {\tt lock\_try\_acquire};
	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait}, {\tt cond\_\allowbreak{}signal}, {\tt cond\_broadcast}, with the same semantics as Pebbles P2 condvars;
	\item Basic round-robin scheduling facilities: {\tt thread\_block} (a kernel-level analogue to Pebbles's {\tt deschedule}), {\tt thread\_yield}
	\item Kernel thread lifecycle management, {\tt thread\_create} and {\tt thread\_exit}, including stack space memory management;
	\item Interrupt and fault handlers;
	\item A page allocator, {\tt palloc\_get\_page}, {\tt palloc\_get\_multiple}, {\tt palloc\_free\_page}, and {\tt palloc\_free\_multiple}
\end{itemize}
Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc}, string-formatting, printing, et cetera.
